---
title: "Writing functions"
description: "How to re-use code while avoiding copy-pasting"
author: 
  - Etienne Bacher
  - Niklas Hohmann
date: "2026-06-18"
format:
  revealjs:
    output-file: index-slides.html
    toc: false
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Functions in R

## Functions in R

<br>

Functions are used everywhere:

- `library()`

- `read.csv()`

- `mean()`

etc.

<br>

. . .

R comes with comes with hundreds of functions by default.

. . . 

Thousands more are available via user-written packages.

## Functions in R

<br>

You can do entire R projects without ever needing to write your own functions.

<br>
<br>

However, knowing how to write a custom function can be extremely useful.

# Demo: standardizing values

## Demo: standardizing values

<br>

We want to [standardize values](https://en.wikipedia.org/wiki/Standard_score) on multiple columns:

$$z = \frac{x - mean(x)}{sd(x)}$$


## Demo: standardizing values

<br>

Option 1: run the code by hand on a list of columns in our data:

<br>

```{r}
iris[["Petal.Length_std"]] <- (iris[["Petal.Length"]] - 
  mean(iris[["Petal.Length"]])) /
  sd(iris[["Petal.Length"]])
iris[["Sepal.Length_std"]] <- (iris[["Sepal.Length"]] - 
  mean(iris[["Sepal.Length"]])) /
  sd(iris[["Sepal.Length"]])
iris[["Petal.Width_std"]] <- (iris[["Petal.Width"]] - 
  mean(iris[["Sepal.Width"]])) /
  sd(iris[["Petal.Width"]])
iris[["Sepal.Width_std"]] <- (iris[["Sepal.Width"]] - 
  mean(iris[["Sepal.Width"]])) /
  sd(iris[["Sepal.Width"]])
```


## Demo: standardizing values

<br>

This works, but what if we want to ignore missing values? We have to add
8 parameters:

```{r}
iris[["Petal.Length_std"]] <- (iris[["Petal.Length"]] - 
  mean(iris[["Petal.Length"]], na.rm = TRUE)) /
  sd(iris[["Petal.Length"]], na.rm = TRUE)
iris[["Sepal.Length_std"]] <- (iris[["Sepal.Length"]] - 
  mean(iris[["Sepal.Length"]], na.rm = TRUE)) /
  sd(iris[["Sepal.Length"]], na.rm = TRUE)
iris[["Petal.Width_std"]] <- (iris[["Petal.Width"]] - 
  mean(iris[["Sepal.Width"]], na.rm = TRUE)) /
  sd(iris[["Petal.Width"]], na.rm = TRUE)
iris[["Sepal.Width_std"]] <- (iris[["Sepal.Width"]] - 
  mean(iris[["Sepal.Width"]], na.rm = TRUE)) /
  sd(iris[["Sepal.Width"]], na.rm = TRUE)
```

. . .

Do you notice anything weird about the code above?


## Demo: standardizing values

<br>

A small typo with big consequences!

```{.r code-line-numbers="8"}
iris[["Petal.Length_std"]] <- (iris[["Petal.Length"]] - 
  mean(iris[["Petal.Length"]], na.rm = TRUE)) /
  sd(iris[["Petal.Length"]], na.rm = TRUE)
iris[["Sepal.Length_std"]] <- (iris[["Sepal.Length"]] - 
  mean(iris[["Sepal.Length"]], na.rm = TRUE)) /
  sd(iris[["Sepal.Length"]], na.rm = TRUE)
iris[["Petal.Width_std"]] <- (iris[["Petal.Width"]] - 
  mean(iris[["Sepal.Width"]], na.rm = TRUE)) /
  sd(iris[["Petal.Width"]], na.rm = TRUE)
iris[["Sepal.Width_std"]] <- (iris[["Sepal.Width"]] - 
  mean(iris[["Sepal.Width"]], na.rm = TRUE)) /
  sd(iris[["Sepal.Width"]], na.rm = TRUE)
```

. . . 

Duplicating code makes it easy to introduce bugs.

Let's write a function instead of repeating this formula four times.

# Writing a function

## The most basic function

<br>

```{r}
my_std <- function() {}
```

<br>

-   `my_std` is the **function name**
-   `function()` is the **function definition**, that can take **function arguments**
-   `{}` contains the **function body**

## The most basic function

<br>

For now, this function is useless:

```{r, eval=TRUE}
my_std()
```

. . .

<br>

Let's add a simple message:

```{r, eval=TRUE}
my_std <- function() {
  print("Hello from my_std()!")
}

my_std()
```

. . .

<br>

Let's add function parameters!

## Introduce function parameters

<br>

Function parameters are inputs passed by the user.

They are the parts that the user can modify to change the output.

<br>

To move existing code into a function, we need to identify its **"moving" components**
and its **"stable" components**.

## Introduce function parameters

<br>

Parts in [red]{style="color:red;"} are the parts that change across lines (the column name that is used):

::: {.sourceCode .cell-code}
```{=html}
<pre class="sourceCode r code-with-copy">
<p><span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #001080;"><span style="color: #ff0000;">"Petal.Length"</span_std><span style="color: #000000;">]] &lt;- (<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #001080;"><span style="color: #ff0000;">"Petal.Length"</span><span style="color: #000000;">]] - <span style="color: #795e26;"><br>  mean<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #001080;"><span style="color: #ff0000;">"Petal.Length"</span><span style="color: #000000;">]])) / <br><span style="color: #795e26;">  sd<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #001080;"><span style="color: #ff0000;">"Petal.Length"</span><span style="color: #000000;">]])</span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span>
<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Length_std"<span style="color: #000000;">]] &lt;- (<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Length"<span style="color: #000000;">]] - <span style="color: #795e26;"><br>  mean<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Length"<span style="color: #000000;">]])) / <br><span style="color: #795e26;">  sd<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Length"<span style="color: #000000;">]])</span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span>
<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Petal.Width_std"<span style="color: #000000;">]] &lt;- (<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Petal.Width"<span style="color: #000000;">]] - <span style="color: #795e26;"><br>  mean<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Petal.Width"<span style="color: #000000;">]])) / <br><span style="color: #795e26;">  sd<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Petal.Width"<span style="color: #000000;">]])</span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span>
<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Width_std"<span style="color: #000000;">]] &lt;- (<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Width"<span style="color: #000000;">]] - <span style="color: #795e26;"><br>  mean<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Width"<span style="color: #000000;">]])) / <br><span style="color: #795e26;">  sd<span style="color: #000000;">(<span style="color: #001080;">iris<span style="color: #000000;">[[<span style="color: #ff0000;">"Sepal.Width"<span style="color: #000000;">]])</span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></span></p>
</pre>
```
:::

## Introduce function parameters

<br>

Therefore we can create a `column` parameter to specify which column to use in the computation:

<br>

```{r}
my_std <- function(column) {}
```

## Add the function body

<br>

The function body takes the code that doesn't change across lines:

```{r}
my_std <- function(column) {
  (iris[["Petal.Length"]] - 
    mean(iris[["Petal.Length"]], na.rm = TRUE)) /
    sd(iris[["Petal.Length"]], na.rm = TRUE)
}
```

. . .

and we can replace the moving parts by our new `column` parameter:

```{r}
my_std <- function(column) {
  (iris[[column]] - mean(iris[[column]], na.rm = TRUE)) /
    sd(iris[[column]], na.rm = TRUE)
}
```

:::{.callout-note} 
We don't assign the output *inside* the function: the function merely makes the computation and we assign its output to our object when we call the function.
:::


## Use our new function

<br>
<br>

```{r}
iris[["Petal.Length_std"]] <- my_std("Petal.Length")
iris[["Sepal.Length_std"]] <- my_std("Sepal.Length")
iris[["Petal.Width_std"]] <- my_std("Petal.Width")
iris[["Sepal.Width_std"]] <- my_std("Sepal.Width")
```

Notice how it is much easier to check whether we made a typo in this code.


## Validate function parameters

<br>

We can pass our own values to `my_std()`, so we can also make mistakes!

If we pass a column name that doesn't exist in the data:

```{r, warning = TRUE}
my_std("unknown_column")
```

. . .

This doesn't error, it just throws a warning and returns a useless result!


{{< fa arrow-right >}} &nbsp; we need to check user inputs


## Validate function parameters

<br>

Step 1: list all input requirements. 

. . .

  - the column should be a single value, e.g.
    `my_std(c("Sepal.Length", "Petal.Width"))` should fail;
  - the column should be a character value, e.g. `my_std(1)` should
    fail;
  - the column should exist in the data, e.g. `my_std("column_1")`
    should fail.

## Validate function parameters

<br>

Step 2: check each of those.

```{.r code-line-numbers="2-4|5-7|8-10|"}
my_std <- function(column) {
  if (length(column) != 1) {
    stop("The `column` parameter must be of length 1.")
  }
  if (!is.character(column)) {
    stop("The `column` parameter must be a character.")
  }
  if (!(column %in% names(iris))) {
    stop("Column '", column, "' doesn't exist in the data.")
  }

  (iris[[column]] - mean(iris[[column]], na.rm = TRUE)) / sd(iris[[column]], na.rm = TRUE)
}
```

```{r, echo=FALSE}
my_std <- function(column) {
  if (length(column) != 1) {
    stop("The `column` parameter must be of length 1.")
  }
  if (!is.character(column)) {
    stop("The `column` parameter must be a character.")
  }
  if (!(column %in% names(iris))) {
    stop("Column '", column, "' doesn't exist in the data.")
  }

  (iris[[column]] - mean(iris[[column]], na.rm = TRUE)) / sd(iris[[column]], na.rm = TRUE)
}
```

## Validate function parameters

<br>

And now we have proper error messages!

```{r, error = TRUE}
my_std(c("Sepal.Length", "Petal.Width"))
my_std(1)
my_std("column_1")
```


## Validate function parameters

::: {.callout-note collapse="true" title="Packages for input checking"}
Several packages provide functions to easily check inputs and report
useful error messages, for example [`rlang`](https://rlang.r-lib.org/reference/index.html#function-arguments) and [`checkmate`](https://mllg.github.io/checkmate/) (see also [`dreamerr`](https://lrberge.github.io/dreamerr/),
[`arg`](https://ngreifer.github.io/arg/),
[`chk`](https://poissonconsulting.github.io/chk/), and more!).

::: {.panel-tabset}

#### With `rlang`

```{r, error = TRUE}
### With rlang
my_std <- function(column) {
  rlang::check_string(column)
  if (!(column %in% names(iris))) {
    stop("Column '", column, "' doesn't exist in the data.")
  }

  (iris[[column]] - mean(iris[[column]], na.rm = TRUE)) / sd(iris[[column]], na.rm = TRUE)
}

my_std(c("Sepal.Length", "Petal.Width"))
my_std(1)
my_std("column_1")
```

#### With `checkmate`

```{r, error = TRUE}
my_std <- function(column) {
  checkmate::assert_string(column)
  checkmate::assert_subset(column, names(iris))

  (iris[[column]] - mean(iris[[column]], na.rm = TRUE)) / sd(iris[[column]], na.rm = TRUE)
}

my_std(c("Sepal.Length", "Petal.Width"))
my_std(1)
my_std("column_1")

```

:::
:::

## Set default value for parameters

<br>

Instead of setting `na.rm = TRUE` in the function body, we could let the user determine this option. 

. . .

We can add a `na.rm` parameter and use its value in `mean()` and `sd()`:

```{r}
my_std <- function(column, na.rm) {
  # [parameter checks]

  (iris[[column]] - mean(iris[[column]], na.rm = na.rm)) / sd(iris[[column]], na.rm = na.rm)
}
```


<br>

<small>For conciseness, I hid the parameter checks we wrote before.</small>


## Set default value for parameters

<br>

This works but forces the user to write `na.rm = TRUE` (or `FALSE`):

```{r}
head(my_std("Sepal.Length", na.rm = FALSE))
```

. . .

We could set a default value instead (like in `mean()` for instance):

```{r}
my_std <- function(column, na.rm = FALSE) {
  # [parameter checks]

  (iris[[column]] - mean(iris[[column]], na.rm = na.rm)) / sd(iris[[column]], na.rm = na.rm)
}
```

```{r}
head(my_std("Sepal.Length"))
```

## Exit a function early

<br>

Sometimes, we want to return an object early on instead of executing the rest of the function.

We can use `return()` for this.

. . .

Example: return `NA` if the input contains at least one `NA` value:

```{r}
f <- function(x) {

  if (anyNA(x)) {
    return(NA)
  }

  # if x doesn't contain NA, then we continue the function here...
}
```

<!-- ## Documenting functions

\[UNSURE: this could fit better in the package development module\]

If you plan to keep using custom functions in your project, or if you
want to share those with colleagues, then it is important to write
documentation. The documentation should include:

-   an explanation of the function's purpose
-   an explanation of each function parameter, including the expected
    type (e.g. "a character value of length 1")
-   details on the expected output -->


# Other ways to use and create functions

## Functions as arguments

<br>

Functions are just another type of objects: they can be updated by the user, assigned to variables, and passed over to other functions:

```{r}
my_variable <- mean
```

<br>

The absence of parentheses after `mean` means we are not evaluating the function, just refering to it.


## Functions as arguments

<br>

The family of `apply` functions can use these function objects:

```{r}
data_list <- list(a = rnorm(100), b = rnorm(100, mean = 3))
lapply(X = data_list, FUN = median)
```


## Anonymous functions

<br>

So far, we have seen and used *named* functions.

Generally, it is good practice to give your variables and functions informative names.

. . .

<br>
But sometimes you just need a quick, short throwaway function
that serves one purpose.

For this, you can use *anonymous* functions. These are defined on the spot and immediately evaluated without being assigned to a variable.

## Anonymous functions

<br>

Once again, these can be very useful in `*apply()` functions (and in other places):

```{r}
data_list <- list("a" = rnorm(100), "b" = rnorm(100, mean = 3))
lapply(X = data_list, FUN = function(x) abs(x[3] - x[5]))
```

## Inspect the source code

<br>

It can be useful to see the inside of a function created by someone else.

Running the function name (without parentheses) in the console shows the source code:
```{r}
my_std
```

## Inspect the source code

::: {.callout-note collapse="true" title="Tip to see the source code of a function"}
In RStudio and Positron, you can press <kbd>Ctrl</kbd> (or <kbd>Cmd</kbd> on macOS) and click on a function name in a script to open a tab containing its source code:

<div style="text-align: center">
![](demo_source_function.webm){width="45%"}
</div>
:::

## Going further

<br>

Designing functions and ensuring their robustness is not trivial. 

There are many choices one can make:

- using positional or named arguments 
- harmonising parameter names
- whether to use default values
- and more

. . .

<br>

Explore more in the [Tidyverse design](https://design.tidyverse.org/) book. 


