---
title: "Workflow design and piping"
description: ""
author: "Etienne Bacher"
date: "2026-07-27"
categories: [r]
difficulty: Beginner
toc: true
format:
  html: default
  revealjs:
    output-file: index-slides.html
execute:
  warning: false
  message: false
  freeze: auto
editor: 
  markdown: 
    wrap: 72
---

# Piping

## Example where we don't use the pipe

Let's say we want to keep the first 10 rows in the `mtcars` data, then keep the observations where `cyl >= 6`, and finally sort the remaining data by the `am` column.

Without using the pipe, we have two main approaches.

First, we can assign each function output to an object.
This object can either keep the same name and be overwritten at each step, or we can use a collection of temporary names:

```{r}
# Same name gets overwritten
res <- head(mtcars, 10)
res <- subset(res, cyl >= 6)
res <- sort_by(res, ~ am)
res

# Temporary names
tmp1 <- head(mtcars, 10)
tmp2 <- subset(tmp1, cyl >= 6)
res <- sort_by(tmp2, ~ am)
res
```

Each of these cases has drawbacks:

- in the first case, if we want to rename `res` in the future then we must be careful to rename all its occurrences throughout the code. It also means that if, say, the call to `sort_by()` is wrong, then we must run the entire block again so that `res` is properly reset. Depending on the data size and operations to run, this can be very time-consuming.

- in the second case, we pollute the global environment with potentially many temporary objects. Additionally, using a counter in a temporary name means that we need to update many names if we want to add an operation between the first and second step for instance.


The second approach isn't to define intermediate objects, but instead to run all those calls at once by nesting functions:

```{r}
res <- sort_by(subset(head(mtcars, 10), cyl >= 6), ~ am)
res

# Same but with different formatting
res <- sort_by(
  subset(
    head(mtcars, 10),
    cyl >= 6
  ),
  ~ am
)
res
```

We didn't define intermediate objects, but to read this code we now need to start from the innermost code (`head(mtcars, 10)`) and expand outwards. 
This may hurt code readability.


## Introducing the pipe

### Using "`|>`"

The pipe is a way to chain operations in natural reading order by automatically passing the output of the left-hand side code to the right-hand side. 
Using the pipe `|>` with the example above would give:

```{r}
res <- head(mtcars, 10) |> 
  subset(cyl >= 6) |> 
  sort_by(~ am)
res
```

The pipe can be read as "and then": we keep the first 10 rows, *and then* we apply our filter, *and then* we sort the remaining data.

We now have some code that doesn't require intermediate objects but is also easy to read because we see operations in the order in which they are executed.

There are a couple of small restrictions in order to use `|>`:

- this operator is available since R 4.1, released in 2021;
- it requires a proper function call in the right-hand side:
  ```{r}
  1:3 |> mean()
  ```
  ```{r}
  #| error: true
  1:3 |> mean
  ```


### Using "`_`"

In the code above, we could seamlessly chain operations because each of those functions take the input data as their first argument.

This is not always the case. For instance, `grepl()` (which detects whether elements of a character vector match a specific pattern) takes the vector to check in second position:

```{r}
let <- letters[1:6]
grepl("a|e", let)
```

Using `|>` without specifying the position of the input would lead to wrong code:

```{r}
#| warning: true
let |> 
  grepl("a|e")
```

The code above is equivalent to `grepl(head(letters), "a|e")`, which is not what we want and is the cause of the warning. 
We want to tell `grepl()` that the data we're passing via `|>` should end up in second position.
To do so, we can use `_`:

```{r}
let |> 
  grepl("a|e", x = _)
```

Note that using `_` comes with a few restrictions:

- this operator is available since R 4.2 (released in 2022);
- `_` must be used on a named argument, e.g. this fails:
  ```{r}
  #| error: true
  let |> 
    grepl("a|e", _)
  ```
- a function call can contain only one `_`:
  ```{r}
  #| error: true
  mtcars$drat |> 
    cor(x = _, y = _)
  ```


### What about "`%>%`"?

You may have seen code that chains operations using `%>%`.
This `%>%` is part of the `magrittr` package.
It predates `|>` and was widely used in the [`tidyverse`](https://tidyverse.org/) until 2021, when `|>` was introduced in base R.

For simple cases, `|>` and `%>%` behave identically.
We recommend using `|>` simply because it is always available in R and doesn't rely on an external package.